---
title: "Government Spending Master DJNF"
author: "Wells" 
date: "2026-05-26"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

Author: Ela Jalil \# Who is getting federal bucks in my state?

For your project, you will examine federal spending in your state and county for the 2020-2024 fiscal years.

We have a dataset that specifies spending for 81 federal agencies and commissions for all 50 states, the District of Columbia and the territories.
You might be able to find a story here… This data comes from [USASpending.gov](http://USASpending.gov) and was compiled by our friends at the Robert H Smith School of Business.

**Here are the drawbacks of this data.**

-   The more detailed we get, the weirder things get.

-   As we get to the county level, some problems arise.
    Some contracts are awarded statewide, so no county is specified.
    So county totals don’t necessarily add up to statewide totals.

-   The state totals will not sum to the overall total.
    That's because some contracts are national and have no state designation.

-   The data is for spending in the United States.
    So the Agency for International Development’s data will be underrepresented here, since they spend money overseas, and ditto for the State Department’s foreign aid programs.

-   The data is for performance in a particular county versus payment to the parent company, which might be in a different state.

### Read the [data dictionary](https://www.usaspending.gov/data-dictionary).

Read the [guide to the data](https://www.usaspending.gov/federal-spending-guide).

Also, worth noting that according to USASpending, the federal obligation variable represents “obligations” rather than actual federal spending.
Actual spending may be lower (a variable represents actual expenses, but it is usually missing or underreported).
In addition, federal obligations recorded in one year can later be corrected through de-obligations in subsequent years.

## Load libraries

```{r include=FALSE}
#install.packages("tidycensus")
#install.packages("janitor")
#install.packages("scales")
library(scales)
library(tidyverse)
library(tidycensus)
library(janitor)
```

## Load spending data

```{r}

spending <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/federal_spending_by_county_2020%202024%20rob_0416_v2.csv") |> 
     clean_names()
#can sort by years, agency, state_nm, cty_name, contracts, grants, direct payments, spending totals
#Check the difference between grants and contracts

```

**pokin' around!**

## Spending by agency, FY2020-2024

```{r}
options(scipen = 999)  # Disable scientific notation globally

totals <- spending |> 
group_by(agency) |> 
  summarize(
    n = n(),
    total_spent = sum(spending_total, na.rm = TRUE), 
    states_spending = sum(spending_total[!is.na(state_nm)], na.rm = TRUE)) |> 
   mutate(`total_spent_$` = scales::dollar(total_spent / 1e9, suffix = "B", accuracy = 0.01)) |> 
  arrange(desc(total_spent))

view(totals)
#is n the amount of times that it's named in the data set?
```

## Spending by state, FY2020-2024

```{r}
state_spending <- spending |> 
group_by(state_nm) |> 
  summarize(
    n = n(),
    total_spent = sum(spending_total, na.rm = TRUE)
  ) |> 
   mutate(`total_spent_$` = scales::dollar(total_spent / 1e9, suffix = "B", accuracy = 0.01)) |> 
  arrange(desc(total_spent))

```

## Spending by county, FY2020-2024

```{r}
county_spending <- spending |> 
group_by(cty_name, state_nm) |> 
  summarize(
     total_spent = sum(spending_total, na.rm = TRUE),
     n = n()
  ) |> 
   mutate(`total_spent_$` = scales::dollar(total_spent / 1e9, suffix = "B", accuracy = 0.01)) |> 
  arrange(desc(total_spent))
#unfortunately a good amount of NA values so may be hard to specify

```

## Highlight specific county, FY2020-2024

```{r}

spending |> 
  filter(cty_name == "HENNEPIN") |> 
  group_by(agency) |> 
  summarize(
     total_spent = sum(spending_total, na.rm = TRUE),
     n = n()
  ) |> 
   mutate(`total_spent_$` = scales::dollar(total_spent / 1e9, suffix = "B", accuracy = 0.01)) |> 
  arrange(desc(total_spent))
```

## Highlight specific city and agency, FY2020-2024

```{r}
spending |> 
  filter(cty_name == "HENNEPIN") |> 
  filter(agency == "Department of Health and Human Services") |> 
   mutate(`total_spent_$` = scales::dollar(total_spent / 1e9, suffix = "B", accuracy = 0.01)) |> 
  select(fyear, cty_name, spending_total)
```

### Follow the Money

To drill down into specific spending by county, we now go to [USASpending.gov](https://www.usaspending.gov/).

-   Select start searching awards
-   Select one fiscal year (multiple years = too much data)\
-   Select location, place of performance and type in the county name
-   Search!

The web page serves up a form with spending by specific company in that county per year.
#Can maybe see if a company has repeatedly shown up **Gold mine!** Make sure this is the search you want.
To download this data, see the download icon in the black toolbar

### Practice queries

Differences across administrations For your state, compare the last year of Trump’s first administration (2020) to the last year of the Biden administration (2021) How did spending differ overall?
Was one administration spending way more/less than another?
Was one administration pouring more/less into specific agencies than the other administration?

```{r}
#AI assistance was used to create this code
#Trump - 2021, or 2020 because it's the last full year
#Biden - 2025, or 2024 because it's the last full year 

# Aggregate total spending per year
spending_by_year <- aggregate(spending_total ~ fyear, data = spending, FUN = sum)

# Sort by year
spending_by_year <- spending_by_year[order(spending_by_year$fyear), ]

# Print the result
print(spending_by_year)

#2020 = 4,246,216,763,125
#2024 = 4,902,206,088,550 - there is almost a 700 billion dollar difference between 2020 and 2024
#Now I'm going to look into which agency recieved the most money under the Trump administration and the Biden administration
trump_final <- spending %>% 
  filter(fyear == 2020) %>% 
  group_by(agency) %>%
  summarise(total_spending = sum(spending_total, na.rm = TRUE)) %>%
  arrange(desc(total_spending))
#Department of Health and Human Services spent the most - 1,560,445,841,005 - I would like to see where this mostly went

biden_final <- spending %>% 
  filter(fyear == 2024) %>% 
  group_by(agency) %>%
  summarise(total_spending = sum(spending_total, na.rm = TRUE)) %>%
  arrange(desc(total_spending))
#Department of Health and Human Services spent the most - could this be explained by COVID? 1,851,928,863,603 - there is a 300 billion dollar increase
#Many of the agencies were the same 
#https://www.usaspending.gov/recipient - was poking around and thought that this was interesting 

```

Creating an API key that can pull the defense contractors and amount in 2020

```{r}
library(httr)
library(jsonlite)
library(dplyr)


body <- list(
  filters = list(
    time_period = list(
      list(
        start_date = "2019-10-01",
        end_date   = "2020-09-30"
      )
    ),
    agencies = list(
      list(
        type = "awarding",
        tier = "toptier",
        name = "Department of Defense"
      )
    )
  ),
  limit = 10,
  page = 1
)

response <- POST(
  "https://api.usaspending.gov/api/v2/search/spending_by_category/recipient/",
  body = body,
  encode = "json"
)

data <- fromJSON(
  content(response, "text", encoding = "UTF-8"),
  flatten = TRUE
)

top10_2020 <- data$results %>%
  select(
    contractor = name,
    uei,
    obligations = amount
  ) %>%
  arrange(desc(obligations))

print(top10_2020)
#Both parent and children corporations were showing up in these results - which I think could still be accurate because they both receive funding. My main concern is just making sure that they are not repeating the amounts. 
#Lockheed Martin received 39,981,831,442 in 2020, and that does not include the child corporation amount
```
Doing the same thing again for 2024
```{r}
library(httr)
library(jsonlite)
library(dplyr)


body <- list(
  filters = list(
    time_period = list(
      list(
        start_date = "2023-10-01",
        end_date   = "2024-09-30"
      )
    ),
    agencies = list(
      list(
        type = "awarding",
        tier = "toptier",
        name = "Department of Defense"
      )
    )
  ),
  limit = 10,
  page = 1
)

response <- POST(
  "https://api.usaspending.gov/api/v2/search/spending_by_category/recipient/",
  body = body,
  encode = "json"
)

data <- fromJSON(
  content(response, "text", encoding = "UTF-8"),
  flatten = TRUE
)

top10_2024 <- data$results %>%
  select(
    contractor = name,
    uei,
    obligations = amount
  ) %>%
  arrange(desc(obligations))

print(top10_2024)
#Both parent and children corporations were showing up in these results - which I think could still be accurate because they both receive funding. My main concern is just making sure that they are not repeating the amounts. 
#Lockheed Martin received 14,357,199,730 - less than half of what they received under the Trump administration 
#Boeing Company was up higher, but still received less money; Humana recieved more money - but that is also the health care for the military 
```


#Could maybe look at something related to organizations that were cut because of DEI?
Need to think of the significant events - what orgs started recieving more money

Aggregate year and state - which contractors in the states were the biggest deal; contractors were
